---
title: "Analysis of GLP1R KO Bone marrow"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: TBA

## Introduction

Here we analyse the effect of transferring GLP1 KO bone marrow into WT mice and vice-versa.
We have diabetic and control groups aswell.

The sample groups are as follows:

* WTWT_CONT: wild type BM transferred into WT recipients, non-diabetic
* WTWT_DIAB: wild type BM transferred into WT recipients, diabetic
* GLPWT_CON: GLP1KO BM transferred into WT recipients, non-diabetic
* GLPWT_DIAB: GLP1KO BM transferred into WT recipients, diabetic
* GLPGLP_CONT: GLP1KO BM transferred into GLP1KO recipients, non-diabetic
* GLPGLP_DIAB: GLP1KO BM transferred into GLP1KO recipients, diabetic
* WTGLP_CONT: wild type BM transferred into GLP1KO recipients, non-diabetic
* WTGLP_DIAB: wild type BM transferred into GLP1KO recipients, diabetic
* WT_CONT: wild type BM, no transfer, non-diabetic
* WT_DIAB: wild type BM, no transfer, diabetic
* GLP_CONT: GLP1KO BM, no transfer, non-diabetic
* GLP_DIAB: GLP1KO BM, no transfer, diabetic

Reads were trimmed using Skewer 0.2.2 (Jiang et al, 2013), discarding bases with quality
scores <20.
Trimmed reads were then mapped to the GENCODE M38 mouse transccriptome (Mudge et al, 2025)
using Kallisto 0.46.2 (Bray et al, 2016).

Genes with mean counts > 10 across all samples in a comparison are classified as detected.

Differential expression is conducted with DESeq2.

Pathway enrichment analysis is conducted with mitch.

Gene sets were obtained from the gene ontology database (q3 2023). Biological process sets were used.

Analysis was conducted in R v4.5.2.

```{r,packages}

suppressPackageStartupMessages({
    library("zoo")
    library("dplyr")
    library("reshape2")
    library("DESeq2")
    library("gplots")
    library("MASS")
    library("mitch")
    library("eulerr")
    library("kableExtra")
    library("beeswarm")
    library("network")

})

knitr::opts_chunk$set(dev = 'svg') # set output device to svg

```

## Import read counts

```{r,importdata1}

tmp <- read.table("3col.tsv.gz",header=F)
x <- as.matrix(acast(tmp, V2~V1, value.var="V3", fun.aggregate = sum))
x <- as.data.frame(x)
accession <- sapply((strsplit(rownames(x),"\\|")),"[[",2)
symbol<-sapply((strsplit(rownames(x),"\\|")),"[[",6)
x$geneid <- paste(accession,symbol)
xx <- aggregate(. ~ geneid,x,sum)
rownames(xx) <- xx$geneid
xx$geneid = NULL
xx <- round(xx)
head(xx)

# remove the version number from the gene IDs
mus_IDs <- sapply(strsplit(rownames(xx),"\\."),"[[",1)
mus_gene_symbols <- sapply(strsplit(rownames(xx)," "),"[[",2)
rownames(xx) <- paste(mus_IDs,mus_gene_symbols,sep=" ")

# keep gene table for later
mus_gt <- data.frame(paste(mus_IDs,mus_gene_symbols,sep=" "),mus_IDs,mus_gene_symbols)
colnames(mus_gt) <- c("IDsymbol","ID","symbol")

```

## Sample sheet

Read in and check the sample sheet.

All datasets are included in the samplesheet.

Not all samples in the sample sheet are in the dataset.

There were 47 samples in the dataset and 57 in the samplesheet.

```{r,ss1}

ss <- read.table("samplesheet.tsv", header=TRUE)

#> #which(ss$sample_name %in% colnames(xx))
which(!colnames(xx) %in% ss$sample_name)

message("Included data")
intersect(ss$sample_name , colnames(xx) )

message("Missing data")
setdiff(ss$sample_name , colnames(xx) )

ss2 <- ss[which(ss$sample_name %in% colnames(xx) ) ,]

ss2 %>% kbl(caption = "Sample sheet") %>% kable_paper("hover", full_width = F)

table(ss2$group)

```

Now I will rename the datasets to make it easier for downstream analysis.
I will also reorder the datasets.

```{r,renamedata}

xx2 <- xx
colnames(xx2) <- ss2[match(colnames(xx),ss2$sample_name),"preferred_samplename"]
ss2 <- ss2[order(ss2$preferred_samplename),]
xx2 <- xx2[,order(colnames(xx2))]

```

## QC analysis

Here I'll look at a few different quality control measures.

```{r,qc1,fig.height=7,fig.width=7}

par(mar=c(5,8,3,1))
barplot(colSums(xx2),horiz=TRUE,las=1,xlab="num reads")
colSums(xx2)

```

We have 277M to 50M assigned reads, which is good.

## MDS plot

We can't see any obvious clustering between samples and there could be two outlers GLPGLP_CONT_3 and WT_CONT_2.

```{r,mds1}

cols <- c(rep("tan1",4),rep("violet",4))

par(mar=c(5.1,4.1,4.1,2.1))

plot(cmdscale(dist(t(xx2))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx2))), labels=colnames(xx2) , cex=0.5)

```

## Correlation heatmap

Correlation heatmap shows a slight batch effect with batch A consisting of replicates 1 and 2,
while batch B consists of replicates 3 and 4.

```{r,cor,fig.height=7,fig.width=7}

heatmap.2(cor(xx2),trace="n",main="Pearson correlation heatmap",cexRow=0.5,cexCol=0.5)

cor(xx2) %>% kbl(caption = "Pearson correlation coefficients") %>% kable_paper("hover", full_width = F)

heatmap.2(cor(xx2,method="s"),trace="n",main="Spearman correlation heatmap",cexRow=0.5,cexCol=0.5)

cor(xx2,method="spearman") %>% kbl(caption = "Spearman correlation coefficients") %>% kable_paper("hover", full_width = F)

```

## Quantification of cell populations

Using data from Nestorowa et al 2016 (A single-cell resolution map of mouse hematopoietic stem and progenitor cell differentiation
).

https://ashpublications.org/blood/article/128/8/e20/35749/A-single-cell-resolution-map-of-mouse
https://blood.stemcells.cam.ac.uk/single_cell_atlas.html

```{r,cellpop1,fig.width=9}

blood <- readRDS("hemoprogenitor.Rds")
genes <- intersect(rownames(xx2), rownames(blood))
dec <- apply(xx2[genes, , drop=F], 2, function(x) coef(rlm( as.matrix(blood[genes,]), x, maxit =100 ))) *100
dec <- t(dec/colSums(dec)*100)
dec <- signif(dec, 3)

# remove negative values
dec2 <- t(apply(dec,2,function(x) { mymin=min(x) ; if (mymin<0) { x + (mymin * -1) } else { x } } ))
dec2 <- apply(dec2,2,function(x) {x / sum(x) *100} )
colfunc <- colorRampPalette(c("blue", "white", "red"))

heatmap.2( dec2, col=colfunc(25),scale="row",
 trace="none",margins = c(7,7), cexRow=.7, cexCol=.6,  main="cell type abundances")

heatmap.2( dec2, col=colfunc(25),scale="none",
 trace="none",margins = c(7,7), cexRow=.7, cexCol=.6,  main="cell type abundances")


heatmap.2( dec2, col=colfunc(25),scale="none",Colv = FALSE,
 trace="none",margins = c(7,7), cexRow=.7, cexCol=.6,  main="cell type abundances")

```

Reorder the heatmap.

```{r,cellpop2,fig.width=9}

TRTGRPS=c("WT_CONT",     "WT_DIAB",
          "GLP_CONT",    "GLP_DIAB",
          "WTWT_CONT",   "WTWT_DIAB",
          "GLPWT_CON",   "GLPWT_DIAB",
          "GLPGLP_CONT", "GLPGLP_DIAB",
          "WTGLP_CONT",  "WTGLP_DIAB")

trtgrp <- paste(sapply(strsplit(colnames(dec2),"_"),"[[",1),sapply(strsplit(colnames(dec2),"_"),"[[",2),sep="_")

dec2l <- lapply(TRTGRPS, function(x) {  dec2[,trtgrp==x] } )
names(dec2l) <- TRTGRPS
dec2o <- do.call(cbind,dec2l)

heatmap.2( dec2o, col=colfunc(25),scale="none",Colv = FALSE,
 trace="none",margins = c(7,7), cexRow=.7, cexCol=.6,  main="cell type abundances")

```

Make some boxplots and compare groups.

```{r,cellpop3}

par(mar=c(7.5,4,3,1))

null <- lapply(1:nrow(dec2o), function(i) {
  CELLNAME <- rownames(dec2o)[i]
  boxplot(lapply(dec2l,function(x) { x[i,]} ),cex=0,col="white",las=2,main=CELLNAME,ylab="Est cell proportion")
  beeswarm(lapply(dec2l,function(x) { x[i,]} ),add=TRUE,cex=2,col="darkgray", pch=19)
})

```

```{r,cellpop4}

par(mar=c(5,10,3,1))
boxplot(t(dec2[order(rowMeans(dec2)),]),horizontal=TRUE,las=1, xlab="estimated cell proportion (%)")
par(mar = c(5.1, 4.1, 4.1, 2.1))
heatmap.2( cor(dec2),trace="none",scale="none", margins = c(7,7))
heatmap.2( cor(t(dec2)),trace="none",scale="none", margins = c(8,8))
par(mar=c(5,10,3,1))
barplot(apply(dec2,1,sd),horiz=TRUE,las=1,xlab="SD of cell proportions (%)")
which(apply(dec2,1,sd)>4)

saveRDS(dec2,"cellcomposition.Rds")

signif(t(dec2),3) %>%
  kbl(caption = "Cell population estimates") %>%
  kable_paper("hover", full_width = F)

sdec2 <- scale(dec2)

ssblood <- t(sdec2[rownames(sdec2) %in% c("Monocytes","Erythroid","Megakaryocytes"),])

ssblood %>%
  kbl(caption = "Cell types that will be adjusted for") %>%
  kable_paper("hover", full_width = F)

ss2 <- cbind(ss2,ssblood)

```

Based on these results, we need to correct for Neutrophils.LD, B.Memory, T.CD4.Naive, Monocytes.C and T.CD4.Memory cells in that order.

## Analysis of differential gene expression

I will set up the following comparisons:

1. WT_CONT vs WT_DIAB: effect of diabetes in WT, no BM transfer
2. GLP_CONT vs GLP_DIAB: effect of diabetes in GLP1RKO, no BM transfer
3. WT_CONT vs GLP_CONT: effect of GLP1RKO in non-diabetic mice
4. WT_DIAB vs GLP_DIAB: effect of GLP1RKO in non-diabetic mice

5. WT_CONT vs WTWT_CONT: effect of the BM transfer process in non-diabetic animals
6. WT_DIAB vs WTWT_DIAB: effect of the BM transfer process in diabetic animals

7. WTWT_DIAB vs GLPWT_DIAB: effect of putting GLP1RKO BM in WT diabetic mice
8. GLPGLP_DIAB vs WTGLP_DIAB: effect of putting WT BM in GLP1RKO diabetic mice

The six possible comparisons of diabetic BM transfer groups.
A. WTWT_DIAB vs GLPWT_DIAB
B. WTWT_DIAB vs GLPGLP_DIAB
C. WTWT_DIAB vs WTGLP_DIAB
D. GLPWT_DIAB vs GLPGLP_DIAB
E. GLPWT_DIAB vs WTGLP_DIAB
F. GLPGLP_DIAB vs WTGLP_DIAB

* WTWT_CONTL: wild type BM transferred into WT recipients, non-diabetic
* WTWT_DIAB: wild type BM transferred into WT recipients, diabetic
* GLPWT_CON: GLP1KO BM transferred into WT recipients, non-diabetic
* GLPWT_DIAB: GLP1KO BM transferred into WT recipients, diabetic
* GLPGLP_CONT: GLP1KO BM transferred into GLP1KO recipients, non-diabetic
* GLPGLP_DIAB: GLP1KO BM transferred into GLP1KO recipients, diabetic
* WTGLP_CONT: wild type BM transferred into GLP1KO recipients, non-diabetic
* WTGLP_DIAB: wild type BM transferred into GLP1KO recipients, diabetic
* WT_CONT: wild type BM, no transfer, non-diabetic
* WT_DIAB: wild type BM, no transfer, diabetic
* GLP_CONT: GLP1KO BM, no transfer, non-diabetic
* GLP_DIAB: GLP1KO BM, no transfer, diabetic

### DGE1 - effect of diabetes in WT, no BM transfer

First we will look at control vs mutant with no filtering for low reads.
Then we remove genes with fewer than 10 reads per sample on average and rerun DESeq2.

```{r,dge1}

ss3 <- subset(ss2,group == "WT_CONT" | group == "WT_DIAB")
ss3$case <- factor(as.numeric(grepl("DIAB",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
#dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by diabetes in WT mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dge1 <- dge

```

#### Make some plots

```{r,plots1}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange, 
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=0.5, col="dark gray",
       main=contrast_name, cex.main=1)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=0.5, col="red")
  mtext(SUBHEADER,cex = 1)
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$pval),cex=0.5,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval")
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$pval),cex=0.5,pch=19,col="red")
}

make_heatmap <- function(de,name,myss,mx,n=30){
  colfunc <- colorRampPalette(c("blue", "white", "red"))
  values <- myss$quickdash
  f <- colorRamp(c("yellow", "orange"))
  rr <- range(values)
  svals <- (values-rr[1])/diff(rr)
  colcols <- rgb(f(svals)/255)
  mxn <- mx/rowSums(mx)*1000000
  x <- mxn[which(rownames(mxn) %in% rownames(head(de,n))),]
  heatmap.2(as.matrix(x),trace="none",col=colfunc(25),scale="row", margins = c(7,15), cexRow=0.9, cexCol=0.9,
    main=paste("Top ranked",n,"genes in",name) )
}

maplot(dge,"ctrl vs diab")
make_volcano(dge,"ctrl vs diab")
make_heatmap(dge,"ctrl vs diab",ss3,xx3f,n=30)

```

### DGE2 - GLP_CONT vs GLP_DIAB: effect of diabetes in GLP1RKO, no BM transfer

```{r,dge2}

ss3 <- subset(ss2,group == "GLP_CONT" | group == "GLP_DIAB")
ss3$case <- factor(as.numeric(grepl("DIAB",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by diabetes in GLP1RKO mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dge2 <- dge

maplot(dge,"ctrl vs diab (GLP1RKO)")
make_volcano(dge,"ctrl vs diab (GLP1RKO)")
make_heatmap(dge,"ctrl vs diab (GLP1RKO)",ss3,xx3f,n=30)

```

### DGE3 - WT_CONT vs GLP_CONT: effect of GLP1RKO in non-diabetic mice

Hmga1b was upregulated, but the most prominent paralog by baseMean expression, Hmga1 was downregulated.
These genes are involved in "regulation of gene transcription, integration of retroviruses into chromosomes,
and the metastatic progression of cancer cells" according to GeneCards.

```{r,dge3}

ss3 <- subset(ss2,group == "WT_CONT" | group == "GLP_CONT")
ss3$case <- factor(as.numeric(grepl("GLP",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by GLPR1KO in non-diabetic mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dge3 <- dge

maplot(dge,"WT vs GLP1RKO (ctrl)")
make_volcano(dge,"WT vs GLP1RKO (ctrl)")
make_heatmap(dge,"WT vs GLP1RKO (ctrl)",ss3,xx3f,n=30)

```

### DGE4 - WT_DIAB vs GLP_DIAB: effect of GLP1RKO in diabetic mice

```{r,dge4}

ss3 <- subset(ss2,group == "WT_DIAB" | group == "GLP_DIAB")
ss3$case <- factor(as.numeric(grepl("GLP",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by GLPR1KO in diabetic mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dge4 <- dge

maplot(dge,"WT vs GLP1RKO (diab)")
make_volcano(dge,"WT vs GLP1RKO (diab)")
make_heatmap(dge,"WT vs GLP1RKO (diab)",ss3,xx3f,n=30)

```

### DGE5 - WT_CONT vs WTWT_CONT: effect of the BM transfer process in non-diabetic animals

```{r,dge5}

ss3 <- subset(ss2,group == "WT_CONT" | group == "WTWT_CONT")
ss3$case <- factor(as.numeric(grepl("WTWT",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by BM transfer in WT non-diabetic mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dge5 <- dge

maplot(dge,"WT vs WTWT (ctrl)")
make_volcano(dge,"WT vs WTWT (ctrl)")
make_heatmap(dge,"WT vs WTWT (ctrl)",ss3,xx3f,n=30)

```


### DGE6 - WT_DIAB vs WTWT_DIAB: effect of the BM transfer process in diabetic animals

```{r,dge6}

ss3 <- subset(ss2,group == "WT_DIAB" | group == "WTWT_DIAB")
ss3$case <- factor(as.numeric(grepl("WTWT",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by BM transfer in WT diabetic mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dge6 <- dge

maplot(dge,"WT vs WTWT (diab)")
make_volcano(dge,"WT vs WTWT (diab)")
make_heatmap(dge,"WT vs WTWT (diab)",ss3,xx3f,n=30)

```


### DGEA - WTWT_DIAB vs GLPWT_DIAB: Effect of transfering GLP1RKO BM into diabetic WT mice

```{r,dgea}

ss3 <- subset(ss2,group == "WTWT_DIAB" | group == "GLPWT_DIAB")
ss3$case <- factor(as.numeric(grepl("GLPWT",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by transfering GLP1RKO BM into non-diabetic WT mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dgea <- dge

maplot(dge,"WTWT vs GLPWT (diab)")
make_volcano(dge,"WTWT vs GLPWT (diab)")
make_heatmap(dge,"WTWT vs GLPWT (diab)",ss3,xx3f,n=30)

```

### DGEB - WTWT_DIAB vs GLPGLP_DIAB: Effect of GLP1RKO in diabetic BM transfer mice

```{r,dgeb}

ss3 <- subset(ss2,group == "WTWT_DIAB" | group == "GLPGLP_DIAB")
ss3$case <- factor(as.numeric(grepl("GLPGLP_DIAB",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by GLP1RKO in diabetic BM transfer mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dgeb <- dge

maplot(dge,"WTWT_DIAB vs GLPGLP_DIAB")
make_volcano(dge,"WTWT_DIAB vs GLPGLP_DIAB")
make_heatmap(dge,"WTWT_DIAB vs GLPGLP_DIAB",ss3,xx3f,n=30)

```

### DGEC - WTWT_DIAB vs WTGLP_DIAB recipient host effect

```{r,dgec}

ss3 <- subset(ss2,group == "WTWT_DIAB" | group == "WTGLP_DIAB")
ss3$case <- factor(as.numeric(grepl("WTGLP_DIAB",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by GLP1RKO host in diabetic BM transfer mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dgec <- dge

maplot(dge,"WTWT_DIAB vs WTGLP_DIAB")
make_volcano(dge,"WTWT_DIAB vs WTGLP_DIAB")
make_heatmap(dge,"WTWT_DIAB vs WTGLP_DIAB",ss3,xx3f,n=30)

```

### DGED - GLPWT_DIAB vs GLPGLP_DIAB donor host effects

```{r,dged}

ss3 <- subset(ss2,group == "GLPWT_DIAB" | group == "GLPGLP_DIAB")
ss3$case <- factor(as.numeric(grepl("GLPGLP_DIAB",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by wt host in diabetic BM transfer mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dged <- dge

maplot(dge,"GLPWT_DIAB vs GLPGLP_DIAB")
make_volcano(dge,"GLPWT_DIAB vs GLPGLP_DIAB")
make_heatmap(dge,"GLPWT_DIAB vs GLPGLP_DIAB",ss3,xx3f,n=30)

```


### DGEE - GLPWT_DIAB vs WTGLP_DIAB 

```{r,dgee}

ss3 <- subset(ss2,group == "GLPWT_DIAB" | group == "WTGLP_DIAB")
ss3$case <- factor(as.numeric(grepl("WTGLP_DIAB",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by wt host in diabetic BM transfer mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dgee <- dge

maplot(dge,"GLPWT_DIAB vs WTGLP_DIAB")
make_volcano(dge,"GLPWT_DIAB vs WTGLP_DIAB")
make_heatmap(dge,"GLPWT_DIAB vs WTGLP_DIAB",ss3,xx3f,n=30)

```


### DGEF - WTGLP_DIAB vs GLPGLP_DIAB

```{r,dgef}

ss3 <- subset(ss2,group == "WTGLP_DIAB" | group == "GLPGLP_DIAB")
ss3$case <- factor(as.numeric(grepl("GLPGLP_DIAB",ss3$group)))

xx3 <- xx2[,which(colnames(xx2) %in% ss3$preferred_samplename)]
xx3f <-  xx3[rowMeans(xx3)>10,]
rpm3 <- apply(xx3f,2,function( x ) { x/sum(x) * 1e6  } )

plot(cmdscale(dist(t(xx3f))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=0 )
text(cmdscale(dist(t(xx3f))), labels=colnames(xx3f) , cex=0.8)

dds <- DESeqDataSetFromMatrix(countData = xx3f , colData = ss3, design = ~ Monocytes + Erythroid + Megakaryocytes + case )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression changes caused by donor GLP1RKO in GLP1RKO diabetic mice") %>%
  kable_paper("hover", full_width = F)

nrow(dge)
nrow(subset(dge,padj<0.05))
dgef <- dge

maplot(dge,"WTGLP_DIAB vs GLPGLP_DIAB")
make_volcano(dge,"WTGLP_DIAB vs GLPGLP_DIAB")
make_heatmap(dge,"WTGLP_DIAB vs GLPGLP_DIAB",ss3,xx3f,n=30)

```

## Pathway enrichment

Here I'm using the mitch package and mouse gene pathways from Gene Ontology Biological Process downloaded from MSigDB
(m5.go.bp.v2025.1.Mm.symbols.gmt).

```{r,gosets}

gobp <- gmt_import("../ref/m5.go.bp.v2025.1.Mm.symbols.gmt")

gt <- as.data.frame(rownames(xx))
gt$gene <- sapply(strsplit(gt[,1]," "),"[[",2)

```

#### DGE1

```{r,m1}

m1 <- mitch_import(dge1, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8,minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top gene pathway differences caused by Diabetes") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="Ctl vs diab (WT)",xlab="ES")
grid()

if ( ! file.exists("mitch_dge1.html") ) {
  mitch_report(res=mr1,outfile="mitch_dge1.html",overwrite=TRUE)
}

```

#### DGE2

```{r,m2}

m1 <- mitch_import(dge2, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8,minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top gene pathway differences caused by diabetes in GLP1KO") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="Ctl vs diab (GLP1RKO)",xlab="ES")
grid()

if ( ! file.exists("mitch_dge2.html") ) {
  mitch_report(res=mr1,outfile="mitch_dge2.html",overwrite=TRUE)
}
mr2 <- mr1

```

#### DGE3

```{r,m3}

m1 <- mitch_import(dge3, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8,minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top gene pathway differences caused by GLP1RKO in non-diabetic mice") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WT vs GLP1RKO (non-diabetic)",xlab="ES")
grid()

if ( ! file.exists("mitch_dge3.html") ) {
  mitch_report(res=mr1,outfile="mitch_dge3.html",overwrite=TRUE)
}
mr3 <- mr1

```

#### DGE4

```{r,m4}

m1 <- mitch_import(dge4, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8,minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top gene pathway differences caused by GLP1RKO in diabetic mice") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WT vs GLP1RKO (diabetic)",xlab="ES")
grid()

if ( ! file.exists("mitch_dge4.html") ) {
  mitch_report(res=mr1,outfile="mitch_dge4.html",overwrite=TRUE)
}
mr4 <- mr1

```

#### DGE5

```{r,m5}

m1 <- mitch_import(dge5, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8,minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top gene pathway differences caused by BM transfer process in non-diabetic animals") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WT_CONT vs WTWT_CONT (non-diabetic)",xlab="ES")
grid()

if ( ! file.exists("mitch_dge5.html") ) {
  mitch_report(res=mr1,outfile="mitch_dge5.html",overwrite=TRUE)
}
mr5 <- mr1

```

#### DGE6

```{r,m6}

m1 <- mitch_import(dge6, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8,minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top gene pathway differences caused by BM transfer process in diabetic animals") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WT_DIAB vs WTWT_DIAB (diabetic)",xlab="ES")
grid()

if ( ! file.exists("mitch_dge6.html") ) {
  mitch_report(res=mr1,outfile="mitch_dge6.html",overwrite=TRUE)
}
mr6 <- mr1

```

#### DGE-A 

```{r,mA}

m1 <- mitch_import(dgea, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8,minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top pathways: Effect of transfering GLP1RKO BM into diabetic WT mice") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WTWT_DIAB vs GLPWT_DIAB",xlab="ES")
grid()

if ( ! file.exists("mitch_dgea.html") ) {
  mitch_report(res=mr1,outfile="mitch_dgea.html",overwrite=TRUE)
}
mra <- mr1

```


#### DGE-B

```{r,mB}

m1 <- mitch_import(dgeb, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8, minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top pathways: Effect of GLP1RKO in diabetic BM transfer mice") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WTWT_DIAB vs GLPGLP_DIAB",xlab="ES")
grid()

if ( ! file.exists("mitch_dgeb.html") ) {
  mitch_report(res=mr1,outfile="mitch_dgeb.html",overwrite=TRUE)
}
mrb <- mr1

```

#### DGE-C 

```{r,mC}

m1 <- mitch_import(dgec, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8, minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top pathways: WTWT_DIAB vs WTGLP_DIAB recipient host effect") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WTWT_DIAB vs WTGLP_DIAB recipient host effect",xlab="ES")
grid()

if ( ! file.exists("mitch_dgec.html") ) {
  mitch_report(res=mr1,outfile="mitch_dgec.html",overwrite=TRUE)
}
mrc <- mr1

```


#### DGE-D

```{r,mD}

m1 <- mitch_import(dged, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8, minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top pathways: GLPWT_DIAB vs GLPGLP_DIAB donor host effects") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="GLPWT_DIAB vs GLPGLP_DIAB donor host effects",xlab="ES")
grid()

if ( ! file.exists("mitch_dged.html") ) {
  mitch_report(res=mr1,outfile="mitch_dged.html",overwrite=TRUE)
}
mrd <- mr1

```


#### DGE-E

```{r,mE}

m1 <- mitch_import(dgee, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8, minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top pathways: GLPWT_DIAB vs WTGLP_DIAB") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="GLPWT_DIAB vs WTGLP_DIAB",xlab="ES")
grid()

if ( ! file.exists("mitch_dgee.html") ) {
  mitch_report(res=mr1,outfile="mitch_dgee.html",overwrite=TRUE)
}
mre <- mr1

```

#### DGE-F

```{r,mF}

m1 <- mitch_import(dgef, DEtype="deseq2",geneTable=gt)
mr1 <- mitch_calc(m1, gobp, priority="effect",cores=8, minsetsize=5)
mres1 <-  subset(mr1$enrichment_result,p.adjustANOVA < 0.01 )
mres1_top <- head(subset(mres1,p.adjustANOVA < 0.05 ),30)
mres1_top <-  mres1_top[order(-mres1_top$s.dist),]

mres1_top %>%
  kbl(caption = "Top pathways: WTGLP_DIAB vs GLPGLP_DIAB") %>%
  kable_paper("hover", full_width = F)

top <- mres1_top
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
par(mar=c(5,27,3,3))
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.65,main="WTGLP_DIAB vs GLPGLP_DIAB",xlab="ES")
grid()

if ( ! file.exists("mitch_dgef.html") ) {
  mitch_report(res=mr1,outfile="mitch_dgef.html",overwrite=TRUE)
}
mrf <- mr1

```


## Presence of marker genes

All charts show log10 transformed data.

Hematopoietic Stem & Progenitor Cells (HSPCs)

HSCs: CD34, CD38 (low), CD90 (Thy1), CD45RA (low), HOPX, MLLT3, AVP
MPPs: CD34, FLT3, CD38 (low)
CMPs/GMPs/MEPs: CD34, CD38, IL3RA, CD135 (FLT3)

Erythroid Lineage

Early: GATA1, KLF1, TAL1
Proerythroblasts → Reticulocytes: GYPA (CD235a), HBA1, HBA2, HBB, ALAS2, EPB42, ANK1
Marker of commitment: TFRC (CD71)

Megakaryocyte Lineage

ITGA2B (CD41), GP1BA (CD42b), PF4, VWF, PPBP, TUBB1, MYH9

Myeloid / Granulocyte Lineage

Granulocyte progenitors: MPO, ELANE, AZU1, PRTN3
Neutrophils: S100A8, S100A9, CSF3R (CD114), FCGR3B (CD16b), CXCR2
Eosinophils: IL5RA, SIGLEC8, CLC, PRG2
Basophils/Mast cells: TPSAB1, CPA3, HDC, MS4A2

Monocyte Lineage

LYZ, CD14, FCGR1A (CD64), CSF1R, S100A8, VCAN, FCN1
Classical mono: CD14 (high), FCGR3A (low)
Non-classical: FCGR3A (CD16, high), CD14 (low), CX3CR1

Dendritic Cells

pDCs: IL3RA (CD123), CLEC4C (BDCA-2), TCF4, LILRA4
cDC1s: CLEC9A, XCR1, CADM1
cDC2s: CD1C, FCER1A, CLEC10A

B Lymphoid Lineage

Pro-B / Pre-B: DNTT (TdT), RAG1, RAG2, VPREB1, IGLL1
Immature/Mature B: CD19, MS4A1 (CD20), PAX5, EBF1, CD79A, CD79B, IGHM
Plasma cells: IGHG1, MZB1, XBP1, PRDM1, SDC1 (CD138)

T / NK Lineage (small populations in marrow)

T cells: CD3D, CD3E, CD3G, TRAC, TRBC
NK cells: NCAM1 (CD56), KLRD1 (CD94), NKG7, GNLY, GZMB, TYROBP
ILCs: IL7R, KLRB1

Stromal / Non-hematopoietic

MSCs: NT5E (CD73), THY1 (CD90), ENG (CD105), CXCL12, LEPR
Endothelial: PECAM1 (CD31), CDH5 (VE-cadherin), KDR, VWF
Osteoblasts: RUNX2, SP7 (Osterix), BGLAP, COL1A1
Adipocytes: ADIPOQ, FABP4, PPARG

```{r,markergenes}

par(mar=c(7.1,4.1,4.1,2.1))

rpm <- apply(xx,2,function(x) { x/sum(x) * 1e6 } )

message("HSC markers: Cd34, Cd38, Thy1, Ptprc, Hopx, Mllt3, Avp")

markers <- c("Cd34$","Cd38$","Thy1$","Ptprc$","Hopx$","Mllt3","Avp$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("MPP markers: CD34, FLT3, CD38 (low)")
markers <- c("Cd34$", "Flt3$", "Cd38$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("CMPs/GMPs/MEPs: CD34, CD38, IL3RA, FLT3")
markers <- c("Cd34$", "Cd38$", "Il3ra$", "Flt3$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Erythroid early lineage")
markers <- c("Gata1$", "Klf1$", "Tal1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Reticulocytes: GYPA (CD235a), HBA1, HBA2, HBB, ALAS2, EPB42, ANK1")
markers <- c("Gypa$","Hba-a1$", "Hbb-y$", "Alas2$", "Epb42$", "Ank1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Marker of commitment: TFRC (CD71)")
markers <- c("Tfrc$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Megakaryocyte Lineage markers: ITGA2B (CD41), GP1BA (CD42b), PF4, VWF, PPBP, TUBB1, MYH9")
markers <- c("Itga2b$","Gp1ba$","Pf4$","Vwf$","Ppbp$","Tubb1$","Myh9$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Myeloid / Granulocyte Lineage")
message("Granulocyte progenitors: MPO, ELANE, AZU1, PRTN3")
markers <- c("Mpo$","Elane$","Tacc2$","Prtn3$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Neutrophils: S100A8, S100A9, CSF3R (CD114), FCGR3B (CD16b), CXCR2")
markers <- c("S100a8$","S100a9$","Csf3r$","Fcgr4$","Cxcr2$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Eosinophils: IL5RA, SIGLEC8, CLC, PRG2")
markers <- c("Il5ra$","Siglece$","Prg2$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Basophils/Mast cells: TPSAB1, CPA3, HDC, MS4A2")
markers <- c("Tpsab1$","Cpa3$","Hdc$","Ms4a2$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Monocyte Lineage: LYZ, CD14, FCGR1A (CD64), CSF1R, S100A8, VCAN, FCN1")
markers <- c("Lyz1$","Lyz2$","Cd14$","Fcgr1$","Csf1r$","S100a8$","Vcan$","Fcnb$","Fcna$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Monocyte Classical mono: CD14 (high), FCGR3A (low)")
markers <- c("Cd14$","Fcgr4")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Monocyte Non-classical: FCGR3A (CD16, high), CD14 (low), CX3CR1")
markers <- c("Fcgr4$","Cd14$","Cx3cr1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Dendritic Cells pDCs: IL3RA (CD123), CLEC4C (BDCA-2), TCF4, LILRA4")
markers <- c("Il3ra$","Clec4b1$","Clec4b2$","Tcf4$","Lilra6$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Dendritic Cells cDC1s: CLEC9A, XCR1, CADM1")
markers <- c("Clec9a$","Xcr1$","Cadm1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Dendritic Cells cDC2s: CD1C, FCER1A, CLEC10A")
markers <- c("Fcer1a$","Clec10a")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("B Lymphoid Lineage Pro-B / Pre-B: DNTT (TdT), RAG1, RAG2, VPREB1, IGLL1")
markers <- c("Dntt$","Rag1$","Rag2$","Vpreb1a$","Vpreb1b$","Igll1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("B Lymphoid Lineage Immature/Mature B: CD19, MS4A1 (CD20), PAX5, EBF1, CD79A, CD79B, IGHM")
markers <- c("Cd19$","Ms4a1$","Ebf1$","Cd79a$","Cd79b$","Ighm$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("B Lymphoid Lineage Plasma cells: IGHG1, MZB1, XBP1, PRDM1, SDC1 (CD138)")
markers <- c("Ighg1$","Mzb1$","Xbp1$","Prdm1$","Sdc1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("T / NK Lineage (small populations in marrow) T cells: CD3D, CD3E, CD3G, TRAC, TRBC")
markers <- c("Cd3d$","Cd3e$","Cd3g$","Trac$","Trbc1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("T / NK Lineage (sml pop) NK cells: NCAM1 (CD56), KLRD1 (CD94), NKG7, GNLY, GZMB, TYROBP")
markers <- c("Ncam1$","Klrd1$","Nkg7$","Tyrobp$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})


message("T / NK Lineage (sml pop) ILCs: IL7R, KLRB1")
markers <- c("Il7r$","Klrb1a$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Stromal / Non-hematopoietic MSCs: NT5E (CD73), THY1 (CD90), ENG (CD105), CXCL12, LEPR")
markers <- c("Nt5e$","Thy1$","Eng$","Cxcl12$","Lepr$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Endothelial: PECAM1 (CD31), CDH5 (VE-cadherin), KDR, VWF")
markers <- c("Pecam1$","Cdh5$","Kdr","Vwf")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Osteoblasts: RUNX2, SP7 (Osterix), BGLAP, COL1A1")
markers <- c("Runx2$","Sp7$","Bglap$","Col1a1$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

message("Adipocytes: ADIPOQ, FABP4, PPARG")
markers <- c("Adipoq$","Fabp4$","Pparg$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

```

Now make some charts for specific genes.

WTWT DIAB VS GLPGLP DIAB:
Il36g ENSMUSG00000044103
Il1b ENSMUSG00000027398

WTWT DIAB VS GLPWT DIAB:
Il9r ENSMUSG00000020279
Nfkbiz ENSMUSG00000035356
Acot1 ENSMUSG00000072949
Prok2 ENSMUSG00000030069
Cxcl2 ENSMUSG00000058427
Hmgcs2 ENSMUSG00000027875

```{r,customcharts1}

par(mar=c(7.1,4.1,4.1,2.1))

message("Genes of interest")
markers <- c("Il36g$","Il1b$", "Il9r$", "Nfkbiz$", "Acot1$", "Prok2$", "Cxcl2$", "Hmgcs2$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

```


WTWT DIAB VS GLPGLP DIAB
and
WTWT DIAB VS GLPWT DIAB

```{r,customcharts2}

par(mar=c(7.1,4.1,4.1,2.1))

message("Genes of interest")

markers <- c("Il36g$","Il1b$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  mygex <- mygex[which(names(mygex) %in% c("WTWT DIAB","GLPGLP DIAB"))]
  #WTWT DIAB VS GLPGLP DIAB
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

markers <- c("Il9r$", "Nfkbiz$", "Acot1$", "Prok2$", "Cxcl2$", "Hmgcs2$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  mygex <- mygex[which(names(mygex) %in% c("WTWT DIAB","GLPWT DIAB"))]
  #WTWT DIAB VS GLPWT DIAB
  boxplot(mygex,log="y",log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

pdf("customcharts.pdf")
par(mar=c(7.1,4.1,4.1,2.1))

message("Genes of interest")

markers <- c("Il36g$","Il1b$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  mygex <- mygex[which(names(mygex) %in% c("WTWT DIAB","GLPGLP DIAB"))]
  #WTWT DIAB VS GLPGLP DIAB
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})

markers <- c("Il9r$", "Nfkbiz$", "Acot1$", "Prok2$", "Cxcl2$", "Hmgcs2$")
null <- lapply(markers,function(markergene) {
  gex <- rpm[grep(markergene,rownames(rpm)),]
  mygex <- lapply(unique(ss$group),function(label) {
    samples <- subset(ss,group==label)$sample_name
    gex[names(gex) %in% samples]
  })
  names(mygex) <- gsub("_"," ",unique(ss$group))
  gene <- gsub("\\$","",markergene)
  mygex <- mygex[which(names(mygex) %in% c("WTWT DIAB","GLPWT DIAB"))]
  #WTWT DIAB VS GLPWT DIAB
  boxplot(mygex,log="y",las=2,col="white",cex=0,main=gene,ylab="reads per million")
  beeswarm(mygex,add=TRUE,cex=2,col="darkgray",pch=19)
})
dev.off()


```

Now make a heatmap and analyse all marker genes.

Values are log10 transformed.

```{r,marker_heatmap,fig.height=13,fig.width=9}

all_marker_genes <- c("Adipoq","Alas2","Ank1","Avp","Bglap","Cadm1","Cd14","Cd19",
  "Cd34","Cd38","Cd3d","Cd3e","Cd3g","Cd79a","Cd79b","Cdh5","Clec10a","Clec4b1",
  "Clec4b2","Clec9a","Col1a1","Cpa3","Csf1r","Csf3r","Cx3cr1","Cxcl12","Cxcr2",
  "Dntt","Ebf1","Elane","Eng","Epb42","Fabp4","Fcer1a","Fcgr1","Fcgr4","Fcna",
  "Fcnb","Flt3","Gata1","Gp1ba","Gypa","Hba-a1","Hbb-y","Hdc","Hopx","Ighg1",
  "Ighm","Igll1","Il3ra","Il5ra","Il7r","Itga2b","Kdr","Klf1","Klrb1a","Klrd1",
  "Lepr","Lilra6","Lyz1","Lyz2","Mllt3","Mpo","Ms4a1","Ms4a2","Myh9","Mzb1",
  "Ncam1","Nkg7","Nt5e","Pecam1","Pf4","Pparg","Ppbp","Prdm1","Prg2","Prtn3",
  "Ptprc","Rag1","Rag2","Runx2","S100a8","S100a9","Sdc1","Siglece","Sp7","Tacc2",
  "Tal1","Tcf4","Tfrc","Thy1","Tpsab1","Trac","Trbc1","Tubb1","Tyrobp","Vcan",
  "Vpreb1a","Vpreb1b","Vwf","Xbp1","Xcr1")

all_marker_genes <- c("Adipoq$","Alas2$","Ank1$","Avp$","Bglap$","Cadm1$","Cd14$","Cd19$",
  "Cd34$","Cd38$","Cd3d$","Cd3e$","Cd3g$","Cd79a$","Cd79b$","Cdh5$","Clec10a$","Clec4b1$",
  "Clec4b2$","Clec9a$","Col1a1$","Cpa3$","Csf1r$","Csf3r$","Cx3cr1$","Cxcl12$","Cxcr2$",
  "Dntt$","Ebf1$","Elane$","Eng$","Epb42$","Fabp4$","Fcer1a$","Fcgr1$","Fcgr4$","Fcna$",
  "Fcnb$","Flt3$","Gata1$","Gp1ba$","Gypa$","Hba-a1$","Hbb-y$","Hdc$","Hopx$","Ighg1$",
  "Ighm$","Igll1$","Il3ra$","Il5ra$","Il7r$","Itga2b$","Kdr$","Klf1$","Klrb1a$","Klrd1$",
  "Lepr$","Lilra6$","Lyz1$","Lyz2$","Mllt3$","Mpo$","Ms4a1$","Ms4a2$","Myh9$","Mzb1$",
  "Ncam1$","Nkg7$","Nt5e$","Pecam1$","Pf4$","Pparg$","Ppbp$","Prdm1$","Prg2$","Prtn3$",
  "Ptprc$","Rag1$","Rag2$","Runx2$","S100a8$","S100a9$","Sdc1$","Siglece$","Sp7$","Tacc2$",
  "Tal1$","Tcf4$","Tfrc$","Thy1$","Tpsab1$","Trac$","Trbc1$","Tubb1$","Tyrobp$","Vcan$",
  "Vpreb1a$","Vpreb1b$","Vwf$","Xbp1$","Xcr1$")

mx <- rpm[unlist(lapply(all_marker_genes,function(x) {grep(x,rownames(rpm)) } )),]
rownames(mx) <- sapply(strsplit(rownames(mx)," "),"[[",2)
colnames(mx) <- ss[match(colnames(rpm),ss$sample_name),"preferred_samplename"]
mx <- mx[,order(colnames(mx))]

heatmap.2( log10(mx+0.1), col=colfunc(25),scale="row",Rowv = FALSE, Colv = FALSE,
 trace="none",margins = c(7,7), cexRow=.5, cexCol=.6,  main="Marker genes")

```

## Session information

```{r,save}

sessionInfo()

save.image("dge2.Rdata")

```
